From Reversible Programs to \\ Univalent Universes and Back

### 2021-10-28 Pi/paper/finite.tex

% What now?
In order to study syntactic descriptions of permutations, we will hit the problem of deciding whether two descriptions
refer to the same permutation -- in group theory, this is the \emph{word problem} for $\Sn$. Putting it in this form
allows us to connect it to the broader scope of computational group theory and combinatorics -- we can borrow ideas such
as Coxeter relations and Lehmer codes.~\footnote{The concepts we use can be found in any standard textbook on group
theory, or see~\cite{symmetryBook2021} for a univalent point of view.}

@@

This construction is called a Coxeter presentation of $\Sn$. Writing it formally, we encode the rules discussed above
using a relation $\cox$ on $\List[\Fin[n]]$ (\cref{def:cox}), and then take its congruence closure $\cox*$
(\cref{def:coxstar}).

@@

Given a permutation $\sigma : \Aut[\Fin[n]]$, for any element $i: \Fin[n]$, we can define the inversion count of $i$ as
the number of smaller elements appearing after it in the permutation.

\begin{definition}[Inversion count]
  Given a permutation $\sigma : \Aut[\Fin[n]]$, the inversion count of $i: \Fin[n]$ is given by
  \[ \invcount{i} = \#\Set{j < i | \sigma(j) > \sigma(i)}. \]
\end{definition}

Just knowing the inversion counts for all the elements, one can reconstruct the starting permutation. Also, observe that
$\invcount{i} < i$, thus fitting in the $i$-th place of a Lehmer code tuple. As an example, consider the following
tabulated presentation of a permutation of $\Fin[5]$.
\[
  \sigma =
  \begin{pmatrix}
    0      & 1      & 2      & 3      & 4      \\
    \el{2} & \el{1} & \el{4} & \el{0} & \el{3} \\
  \end{pmatrix}
\]

@@

\noindent The Lehmer code for the permutation $\sigma$ is then the 5-tuple
\[
  l = (\invcount{0}, \invcount{1}, \invcount{2}, \invcount{3}, \invcount{4}) = (0, 1, 2, 0, 2)
\]

@@

Finally, it is time to complete our goal of characterising the permutation groups. Having produced a Lehmer code by
normalising words in $\Sn$, we need to run it to produce a concrete bijection of finite sets, and, given a bijection
between finite sets, we need to encode it as a Lehmer code. We will prove that these maps construct an equivalence
between the types $\Lehmer[n]$ and $\Aut[\Fin[\suc[n]]]$. The idea for this proof is borrowed
from~\cite{Molzer-cubical}. ~\footnote{Note that the indices for the type of permutations are off-by-one, because we
  chose $\Fin[n]$ to represent generators for permutations on $\Fin[\suc[n]]$.}

@@

\begin{proof}
  For $n = \zero$, note that $\Lehmer[\zero]$ is contractible, and so is $\Aut[\Fin[\suc[\zero]]]$. For $n = \suc[m]$,
  we compute a chain of equivalences.
  \begin{gather*}
    \arraycolsep=0.5em\def\arraystretch{1.5}
    %% \begin{array}{rl}
    %%        & \Lehmer[\zero]          \\
    %%   \eqv & \unit                   \\
    %%   \eqv & \Aut[\Fin[\suc[\zero]]]
    %% \end{array}
    %% \qquad\qquad
    \begin{array}{rlr}
           & \Lehmer[\suc[m]]                                                                                    &                                                                 \\
      \eqv & \Fin[\suc[\suc[m]]] \times \Lehmer[m]                                                               & \text{by definition}                                            \\
      \eqv & \Fin[\suc[\suc[m]]] \times \Aut[\Fin[\suc[m]]]                                                      & \text{induction hypothesis}                                     \\
      \eqv & \dsum*{k : \Fin[\suc[\suc[m]]]}{\Fin[\suc[m]] \eqv \Fin[\suc[m]]}                                   & \text{$\Sigma$ over a constant family}                          \\
      \eqv & \dsum*{k : \Fin[\suc[\suc[m]]]}{\FinExcept{\suc[\suc[m]]}[m] \eqv \Fin[\suc[m]]}                    & \text{by~\cref{prop:fin-finexcept}~\cref{prop:fin-finexcept-2}} \\
      \eqv & \dsum*{k : \Fin[\suc[\suc[m]]]}{\FinExcept{\suc[\suc[m]]}[m] \eqv \FinExcept{\suc[\suc[m]]}[m - k]} & \text{by~\cref{prop:fin-finexcept}~\cref{prop:fin-finexcept-2}} \\
      \eqv & \Aut[\Fin[\suc[\suc[m]]]]                                                                           & \text{by~\cref{prop:fin-finexcept}~\cref{prop:fin-finexcept-3}}
    \end{array}
  \end{gather*}
\end{proof}

>>>>

% What now?
In order to study syntactic descriptions of permutations, we will hit the problem of deciding whether two descriptions
refer to the same permutation -- in group theory, this is the \emph{word problem} for $\Sn$. Putting it in this form
allows us to connect it to the broader scope of computational group theory and combinatorics -- we can borrow ideas such
as Coxeter relations and Lehmer codes. The concepts we use can be found in any standard textbook on group theory, or see
the Symmetry book~\cite{symmetryBook2021} for a univalent point of view.

@@

\noindent
This construction is called a Coxeter presentation of $\Sn$. Writing it formally, we encode the rules discussed above
using a relation $\cox$ on $\List[\Fin[n]]$ (\cref{def:cox}), and then take its congruence closure $\cox*$
(\cref{def:coxstar}).

@@

The inversion count of an element is the number of smaller elements appearing after it in the permutation.

\begin{definition}[Inversion count]
  Given a permutation $\sigma : \Aut[\Fin[n]]$, the inversion count of an element $i: \Fin[n]$ is given by
  \[ \invcount{i} \defeq \#\Set{j < i | \sigma(j) > \sigma(i)}. \]
\end{definition}

Knowing the inversion counts for all the elements, one can reconstruct the starting permutation. Also, observe that
$\invcount{i} < i$, thus fitting in the $i$-th place of a Lehmer code tuple. As an example, consider the following
tabulated presentation of a permutation of $\Fin[5]$.
\[
  \sigma \defeq
  \begin{pmatrix}
    0      & 1      & 2      & 3      & 4      \\
    \el{2} & \el{1} & \el{4} & \el{0} & \el{3} \\
  \end{pmatrix}
\]

@@

\noindent The Lehmer code for the permutation $\sigma$ is then the 5-tuple
\[
  l \equiv (\invcount{0}, \invcount{1}, \invcount{2}, \invcount{3}, \invcount{4}) \equiv (0, 1, 2, 0, 2)
\]

@@

Finally, it is time to complete our goal of characterising the permutation groups. Having produced a Lehmer code by
normalising words in $\Sn$, we need to run it to produce a concrete bijection of finite sets, and, given a bijection
between finite sets, we need to encode it as a Lehmer code. We will prove that these maps construct an equivalence
between the types $\Lehmer[n]$ and $\Aut[\Fin[\suc[n]]]$~\footnote{Note that the indices for the type of permutations
  are off-by-one, because we chose $\Fin[n]$ to represent generators for permutations on $\Fin[\suc[n]]$.}. The idea for
this proof is borrowed from~\citet{Molzer-cubical}.

@@

\begin{proof}
  For $n = \zero$, note that $\Lehmer[\zero]$ is contractible, and so is $\Aut[\Fin[\suc[\zero]]]$, hence we have that
  ${\Lehmer[\zero] \eqv \unit \eqv \Aut[\Fin[\suc[\zero]]]}$. For $n = \suc[m]$, we compute the following chain of equivalences.
  \[
    \arraycolsep=0.5em\def\arraystretch{1.3}
    \begin{array}{rlr}
           & \Lehmer[\suc[m]]                                                                                    &                                                                     \\
      \eqv & \Fin[\suc[\suc[m]]] \times \Lehmer[m]                                                               & \text{by definition}                                                \\
      \eqv & \Fin[\suc[\suc[m]]] \times \Aut[\Fin[\suc[m]]]                                                      & \text{induction hypothesis}                                         \\
      \eqv & \dsum*{k : \Fin[\suc[\suc[m]]]}{\Fin[\suc[m]] \eqv \Fin[\suc[m]]}                                   & \text{$\Sigma$ over a constant family}                              \\
      \eqv & \dsum*{k : \Fin[\suc[\suc[m]]]}{\FinExcept{\suc[\suc[m]]}[m] \eqv \Fin[\suc[m]]}                    & \text{by~\cref{prop:fin-finexcept}~part~\ref{prop:fin-finexcept-2}} \\
      \eqv & \dsum*{k : \Fin[\suc[\suc[m]]]}{\FinExcept{\suc[\suc[m]]}[m] \eqv \FinExcept{\suc[\suc[m]]}[m - k]} & \text{by~\cref{prop:fin-finexcept}~part~\ref{prop:fin-finexcept-2}} \\
      \eqv & \Aut[\Fin[\suc[\suc[m]]]]                                                                           & \text{by~\cref{prop:fin-finexcept}~part~\ref{prop:fin-finexcept-3}}
    \end{array}
  \]
\end{proof}

### 2021-10-28 Pi/paper/introduction.tex

      \item ~\cref{sec:ufin} describes the construction of $\UFin$, the groupoid of finite sets and permutations, in
            Homotopy Type Theory. We define and characterise the notion of a univalent subuniverse, and construct

@@

            formalisation. Our results are stated using HoTT~\cite{univalentfoundationsprogramHomotopyTypeTheory2013},

@@

\noindent
The proofs of some of our lemmas and propositions and theorems, as well as additional material, are given in the
supplementary appendices in the extended version~\cite{choudhurySymmetriesReversibleProgramming2021}, and we refer to
them in the text. Our accompanying Agda code~\cite{choudhuryArtifactSymmetriesReversible2021} contains the formalisation
of the full syntax, and most of our proofs.

>>>>

      \item \cref{sec:ufin} describes the construction of $\UFin$, the groupoid of finite sets and permutations, in
            Homotopy Type Theory and Univalent Foundations. We define and characterise the notion of a univalent subuniverse, and construct

@@

            formalisation. Our results are stated using HoTT/UF~\cite{univalentfoundationsprogramHomotopyTypeTheory2013},

@@

\noindent
The proofs of some of our lemmas and propositions and theorems, as well as additional material, are given in the
supplementary appendices in the extended version~\cite*{choudhurySymmetriesReversibleProgramming2021}, and we refer to
them in the text. Our accompanying Agda code~\cite*{choudhuryArtifactSymmetriesReversible2021} contains the
formalisation of the full syntax, and most of our proofs.

### 2021-10-28 Pi/paper/reversible1.tex
Use uniform style for references
Move macros to syntax.sty

\begin{figure}[t]
  {\scalebox{\scalef}{$%
        %%\noindent\begin{minipage}{.7\linewidth}
        \begin{array}{rrcll}
          \idc :     & A                     & \isoone & A                            & : \idc      \\
          \\
          \identlp : & \zerot + A            & \isoone & A                            & : \identrp  \\
          \swapp :   & A + B                 & \isoone & B + A                        & : \swapp    \\
          \assoclp : & A + (B + C)           & \isoone & (A + B) + C                  & : \assocrp  \\ [1.5ex]
          \identlt : & \onet \times A        & \isoone & A                            & : \identrt  \\
          \swapt :   & A \times B            & \isoone & B \times A                   & : \swapt    \\
          \assoclt : & A \times (B \times C) & \isoone & (A \times B) \times C        & : \assocrt  \\ [1.5ex]
          \absorbr : & ~ \zerot \times A     & \isoone & \zerot ~                     & : \factorzl \\
          \dist :    & ~ (A + B) \times C    & \isoone & (A \times C) + (B \times C)~ & : \factor
        \end{array}$}}

@@

In reversible boolean circuits, the number of input bits matches the number of output bits. Thus, a key insight for a
programming language of reversible circuits is to ensure that each primitive operation preserves the number of bits,
which is just a natural number. The algebraic structure of natural numbers as the free commutative semiring (or,
commutative rig), with $(0,+)$ for addition, and $(1,\times)$ for multiplication then provides sequential, vertical, and
horizontal circuit composition operators.

These commutative rig identities can be used to design a logic for reversible
programming~\cite*{sparksSuperstructuralReversibleLogic2014}. To interpret natural number identities as reversible programs,
the logic needs to be equipped with values and types, and a notion of operational semantics and contextual equivalence,
giving a computational interpretation of the commutative rig structure~\cite{jamesInformationEffects2012}. On the
semantic side, the natural space to consider is the groupoidification of a commutative rig, that is, a symmetric rig
groupoid.

% In programming parlance, as shown in~\Cref{fig:pi-terms}, each primitive algebraic identity of commutative rigs becomes
% a reversible combinator in the programming language and the algebraic closure operators become composition operators in
% the programming language. Once equipped with a notion of values and types, we get the following syntactic presentation
% of a programming language for permutations~\cite{James:2012:IE:2103656.2103667,Carette2016}:

{\scalebox{\scalef}{$%
      \begin{array}{lrcl}
        \textit{Value types}   & A,B,C,D & ::= & \zerot \alt \onet \alt A+B \alt A\times B        \\
        \textit{Values}        & v,w,x,y & ::= & \Acon{tt} \alt \inlv{v} \alt \inrv{v} \alt (v,w) \\
        \textit{Program types} &         &     & A \isoone B                                      \\
        \textit{Programs}      & c       & ::= & (\textrm{See Fig.~\ref{fig:pi-terms}})
      \end{array}$}}

@@

To see how this language expresses reversible circuits, we present a few examples in the Agda embedding of the language.
First it is possible to directly mimic the \verb|qasm|-perspective by defining types that describe sequences of
booleans. We use the type $\mathbb{2} = \onet + \onet$ to represent booleans with $\inlv{\Acon{tt}}$ representing
\textsf{true} and $\inrv{\Acon{tt}}$ representing $\textsf{false}$. Boolean negation (the \verb|x|-gate) is
straightforward to define using the primitive combinator $\swapp$. We can represent $n$-bit words using an n-ary product
of boolean values, thus the type $\mathbb{2} \times (\mathbb{2} \times \mathbb{2})$ (abbreviated $\mathbb{B}~3$)
corresponds to a collection of wires that can transmit three bits.
%
% For example, we can express a 3-bit word reversal operation as follows:
% $\Afun{reverse} : \mathbb{B}~3 \iso \mathbb{B}~3$
% $\Afun{reverse} = \swapt \circledcirc (\swapt  \otimes  \idc)~ \circledcirc \assocrt$
% \noindent The manual trace of $\Afun{reverse}$ below confirms that it indeed reverses the three bits:
% \[\begin{array}{rlr}
%                          & (v_1, (v_2, v_3)) \\
%     \swapt               & ((v_2, v_3), v_1) \\
%     \swapt \otimes  \idc & ((v_3, v_2), v_1) \\
%     \assocrt             & (v_3, (v_2, v_1)) \\
%   \end{array}\]
%subcode source isomorphisms.tex:979
%
To express the \verb|cx| and \verb|ccx| gates, we need to encode a notion of conditional expression. Such conditionals
turn out to be expressible using the distributivity and factoring identities of rigs as shown below:

@@

\noindent where we clearly see the sequences of the three operations \verb|ccx|, \verb|cx|, and \verb|cx| but, instead
of using the indices in the sequence of wires to identify the relevant parameters, here we use structural isomorphisms
to re-shuffle the types. We only show one of these re-shuffling isomorphisms and elide the others. For the second circuit, instead of transcribing it directly, we express it using
a slightly more abstract notation:

@@

All of this is only half the story, however. A sound semantics for $\PiLang$ in weak rig
groupoids was established by \citet{caretteComputingSemiringsWeak2016}, and conjectured to be complete. For this
semantics, \emph{coherence conditions} for symmetric rig groupoids that identify different syntactic representations of the same
permutation~\cite{laplaza72,caretteComputingSemiringsWeak2016}, were collected  in a
second level of $\PiLang$ syntax as level-2 combinators.  Each level-2 combinator is of the form $c_1 \isotwo c_2$ for appropriate
$c_1$ and $c_2$ of the same level-1 type $A \isoone B$ and asserts that $c_1$ and $c_2$ denote the same bijection. For
example, we have the following level-2 combinators dealing with associativity:

@@

\begin{multicols}{2}
  %%  \begin{longtable}{>{$}r<{$} >{$}l<{$} >{$}c<{$} >{$}l<{$}}
  {\scalebox{\scalef}{$%
        \begin{tabular}[t]{>{$}r<{$} @{\hspace{4pt}} >{$}l<{$} >{$}c<{$} >{$}l<{$}}
          \pdenot{\identlp} & (\inl{v})         & = & v               \\
          \pdenot{\identrp} & v                 & = & \inl{v}         \\
          \pdenot{\swapp}   & (\inl{v})         & = & \inr{v}         \\
          \pdenot{\swapp}   & (\inr{v})         & = & \inl{v}         \\
          \pdenot{\assoclp} & (\inl{v})         & = & \inl{(\inl{v})} \\
          \pdenot{\assoclp} & (\inr{(\inl{v})}) & = & \inl{(\inr{v})} \\
          \pdenot{\assoclp} & (\inr{(\inr{v})}) & = & \inr{v}         \\
          \pdenot{\assocrp} & (\inl{(\inl{v})}) & = & \inl{v}         \\
          \pdenot{\assocrp} & (\inl{(\inr{v})}) & = & \inr{(\inl{v})} \\
          \pdenot{\assocrp} & (\inr{v})         & = & \inr{(\inr{v})}
        \end{tabular}$}}

  {\scalebox{\scalef}{$%
        \begin{tabular}[t]{>{$}r<{$} @{\hspace{4pt}} >{$}l<{$} >{$}c<{$} >{$}l<{$}}
          \pdenot{\identlt} & (\ttt , v)          & = & v                   \\
          \pdenot{\identrt} & v                   & = & (\ttt , v)          \\
          \pdenot{\swapt}   & (v_1 , v_2)         & = & (v_2 , v_1)         \\
          \pdenot{\assoclt} & (v_1 , (v_2 , v_3)) & = & ((v_1 , v_2) , v_3) \\
          \pdenot{\assocrt} & ((v_1 , v_2) , v_3) & = & (v_1 , (v_2 , v_3)) \\
          \pdenot{\dist}    & (\inl{v_1} , v_3)   & = & \inl{(v_1 , v_3)}   \\
          \pdenot{\dist}    & (\inr{v_2 , v_3})   & = & \inr{(v_2 , v_3)}   \\
          \pdenot{\factor}  & (\inl{(v_1 , v_3)}) & = & (\inl{v_1} , v_3)   \\
          \pdenot{\factor}  & (\inr{(v_2 , v_3)}) & = & (\inr{v_2} , v_3)   \\
          \pdenot{\idc}     & v                   & = & v
        \end{tabular}$}}
  %%\end{longtable}
\end{multicols}

\begin{center}
  {\scalebox{\scalef}{$%
        \begin{tabular}[t]{>{$}r<{$} @{\hspace{4pt}} >{$}l<{$} >{$}c<{$} >{$}l<{$}}
          \pdenot{(c_1 \circledcirc c_2)} & v           & = & \denot{c_2} (\denot{c_1}~v)         \\
          \pdenot{(c_1 \oplus c_2)}       & (\inl{v})   & = & \inl{(\denot{c_1}~v)}               \\
          \pdenot{(c_1 \oplus c_2)}       & (\inr{v})   & = & \inr{(\denot{c_2}~v)}               \\
          \pdenot{(c_1 \otimes c_2)}      & (v_1 , v_2) & = & (\denot{c_1}~v_1 , \denot{c_2}~v_2)
        \end{tabular}
      $}}
\end{center}

@@

\begin{theoremrep}\label{thm:semone}
  The semantics is sound in the following sense:
  \begin{itemize}
    \item For every level-1 combinator $c : A \isoone B$, we have a bijection $\denot{c}$ between $\denot{A}$ and $\denot{B}$.
    \item For every pair of combinators $c_1$ and $c_2$ of the same type $A \isoone B$, if there exists a level-2
          combinator $\alpha : c_1 \isotwo c_2$, then $\denot{c_1} = \denot{c_2}$ using extensional equivalence of
          functions.
  \end{itemize}
\end{theoremrep}
\begin{proof}
  For every primitive combinator $c$ listed on one side of Fig.~\ref{fig:pi-terms}, let $!c$ be the combinator listed on
  the other side. Thus $! \assoclp$ is $\assocrp$ and $! \swapp$ is $\swapp$ itself. Then we have that $\denot{c}$ and
  $\denot{!c}$ form an equivalence. For the level-2 combinator \Afun{idr◎l}, we check
  $\denot{\AgdaBound{c}~\AgdaOperator{\AgdaInductiveConstructor{◎}}~\AgdaInductiveConstructor{id⟷₁}}
    = \mathit{id} \circ \denot{c} = \denot{c}$. The other cases are only slightly more involved calculations.
\end{proof}

>>>>

In reversible boolean circuits, the number of input bits matches the number of output bits. Thus, a key insight for a
programming language of reversible circuits is to ensure that each primitive operation preserves the number of bits,
which is just a natural number. The algebraic structure of natural numbers as the free commutative semiring (or,
commutative rig), with $(0,+)$ for addition, and $(1,\times)$ for multiplication then provides sequential, vertical, and
horizontal circuit composition.

These commutative rig identities can be used to design a logic for reversible
programming~\cite*{jamesInformationEffects2012,sparksSuperstructuralReversibleLogic2014}. To interpret natural number
identities as reversible programs, the logic needs to be equipped with values and types, and a notion of operational
semantics and contextual equivalence, giving a computational interpretation of the commutative rig
structure~\cite{jamesInformationEffects2012}. On the semantic side, the natural space to consider is the
groupoidification of a commutative rig, that is, a symmetric rig groupoid.

% In programming parlance, as shown in~\cref{fig:pi-terms}, each primitive algebraic identity of commutative rigs becomes
% a reversible combinator in the programming language and the algebraic closure operators become composition operators in
% the programming language. Once equipped with a notion of values and types, we get the following syntactic presentation
% of a programming language for permutations~\cite{James:2012:IE:2103656.2103667,Carette2016}:

{\scalebox{\scalef}{$%
      \begin{array}{lrcl}
        \textit{Value types}   & A,B,C,D & ::= & \zerot \alt \onet \alt A+B \alt A\times B        \\
        \textit{Values}        & v,w,x,y & ::= & \Acon{tt} \alt \inlv{v} \alt \inrv{v} \alt (v,w) \\
        \textit{Program types} &         &     & A \isoone B                                      \\
        \textit{Programs}      & c       & ::= & (\textrm{See~\cref{fig:pi-terms}})
      \end{array}$}}

\begin{figure}
  {\scalebox{\scalef}{$%
        %%\noindent\begin{minipage}{.7\linewidth}
        \begin{array}{rrcll}
          \idc :     & A                     & \isoone & A                            & : \idc      \\ [1.5ex]
          \identlp : & \zerot + A            & \isoone & A                            & : \identrp  \\
          \swapp :   & A + B                 & \isoone & B + A                        & : \swapp    \\
          \assoclp : & A + (B + C)           & \isoone & (A + B) + C                  & : \assocrp  \\ [1.5ex]
          \identlt : & \onet \times A        & \isoone & A                            & : \identrt  \\
          \swapt :   & A \times B            & \isoone & B \times A                   & : \swapt    \\
          \assoclt : & A \times (B \times C) & \isoone & (A \times B) \times C        & : \assocrt  \\ [1.5ex]
          \absorbr : & ~ \zerot \times A     & \isoone & \zerot ~                     & : \factorzl \\
          \dist :    & ~ (A + B) \times C    & \isoone & (A \times C) + (B \times C)~ & : \factor
        \end{array}$}}

@@

@@

To see how this language expresses reversible circuits, we present a few examples using the Agda embedding of the
language~\cite{choudhuryArtifactSymmetriesReversible2021}. First it is possible to directly mimic the
\verb|qasm|-perspective by defining types that describe sequences of booleans ($\mathbb{2}^{n}$). We use the type
$\mathbb{2} = \onet + \onet$ to represent booleans with $\inlv{\Acon{tt}}$ representing \textsf{true}, and
$\inrv{\Acon{tt}}$ representing $\textsf{false}$. Boolean negation (the \verb|x|-gate) is straightforward to define
using the primitive combinator $\swapp$. We can represent $n$-bit words using an n-ary product of boolean values, thus
the type $\mathbb{2} \times (\mathbb{2} \times \mathbb{2})$ (abbreviated $\mathbb{B}~3$) corresponds to a collection of
wires that can transmit three bits.
%
% For example, we can express a 3-bit word reversal operation as follows:
% $\Afun{reverse} : \mathbb{B}~3 \iso \mathbb{B}~3$
% $\Afun{reverse} = \swapt \circledcirc (\swapt  \otimes  \idc)~ \circledcirc \assocrt$
% \noindent The manual trace of $\Afun{reverse}$ below confirms that it indeed reverses the three bits:
% \[\begin{array}{rlr}
%                          & (v_1, (v_2, v_3)) \\
%     \swapt               & ((v_2, v_3), v_1) \\
%     \swapt \otimes  \idc & ((v_3, v_2), v_1) \\
%     \assocrt             & (v_3, (v_2, v_1)) \\
%   \end{array}\]
%subcode source isomorphisms.tex:979
%
To express the \verb|cx| and \verb|ccx| gates, we need to encode a notion of conditional expression. Such conditionals
turn out to be expressible using the distributivity and factoring identities of rigs as shown below:

@@

\noindent where we clearly see the sequences of the three operations \verb|ccx|, \verb|cx|, and \verb|cx| but, instead
of using the indices in the sequence of wires to identify the relevant parameters, here we use structural isomorphisms
to re-shuffle the types. We only show one of these re-shuffling isomorphisms and elide the others. For the second
circuit, instead of transcribing it directly, we express it using $\Afun{cif}$.

@@

All of this is only half the story, however. A sound semantics for $\PiLang$ using weak rig
groupoids was established by~\citet{caretteComputingSemiringsWeak2016}, and conjectured to be complete. For this
semantics, \emph{coherence conditions} for symmetric rig groupoids that identify different syntactic representations of the same
permutation~\cite{laplaza72,caretteComputingSemiringsWeak2016}, were collected  in a
second level of $\PiLang$ syntax as level-2 combinators.  Each level-2 combinator is of the form $c_1 \isotwo c_2$ for appropriate
$c_1$ and $c_2$ of the same level-1 type $A \isoone B$ and asserts that $c_1$ and $c_2$ denote the same bijection. For
example, we have the following level-2 combinators dealing with associativity:

@@

\begin{center}
\begin{tabular}[t]{c @{\hspace{5em}} c}
  {\scalebox{\scalef}{
      \begin{tabular}[t]{>{$}r<{$} @{\hspace{4pt}} >{$}l<{$} >{$}c<{$} >{$}l<{$}}
        \denotbox{\identlp} & (\inl{v})         & = & v               \\
        \denotbox{\identrp} & v                 & = & \inl{v}         \\
        \denotbox{\swapp}   & (\inl{v})         & = & \inr{v}         \\
        \denotbox{\swapp}   & (\inr{v})         & = & \inl{v}         \\
        \denotbox{\assoclp} & (\inl{v})         & = & \inl{(\inl{v})} \\
        \denotbox{\assoclp} & (\inr{(\inl{v})}) & = & \inl{(\inr{v})} \\
        \denotbox{\assoclp} & (\inr{(\inr{v})}) & = & \inr{v}         \\
        \denotbox{\assocrp} & (\inl{(\inl{v})}) & = & \inl{v}         \\
        \denotbox{\assocrp} & (\inl{(\inr{v})}) & = & \inr{(\inl{v})} \\
        \denotbox{\assocrp} & (\inr{v})         & = & \inr{(\inr{v})}
      \end{tabular}}}
  &
  {\scalebox{\scalef}{
      \begin{tabular}[t]{>{$}r<{$} @{\hspace{4pt}} >{$}l<{$} >{$}c<{$} >{$}l<{$}}
        \denotbox{\identlt} & (\ttt , v)          & = & v                   \\
        \denotbox{\identrt} & v                   & = & (\ttt , v)          \\
        \denotbox{\swapt}   & (v_1 , v_2)         & = & (v_2 , v_1)         \\
        \denotbox{\assoclt} & (v_1 , (v_2 , v_3)) & = & ((v_1 , v_2) , v_3) \\
        \denotbox{\assocrt} & ((v_1 , v_2) , v_3) & = & (v_1 , (v_2 , v_3)) \\
        \denotbox{\dist}    & (\inl{v_1} , v_3)   & = & \inl{(v_1 , v_3)}   \\
        \denotbox{\dist}    & (\inr{v_2 , v_3})   & = & \inr{(v_2 , v_3)}   \\
        \denotbox{\factor}  & (\inl{(v_1 , v_3)}) & = & (\inl{v_1} , v_3)   \\
        \denotbox{\factor}  & (\inr{(v_2 , v_3)}) & = & (\inr{v_2} , v_3)   \\
        \denotbox{\idc}     & v                   & = & v
      \end{tabular}}}
\end{tabular}
\end{center}

\begin{center}
  {\scalebox{\scalef}{
      \begin{tabular}[t]{>{$}r<{$} @{\hspace{4pt}} >{$}l<{$} >{$}c<{$} >{$}l<{$}}
        \denotbox{(c_1 \circledcirc c_2)} & v           & = & \denot{c_2} (\denot{c_1}~v)         \\
        \denotbox{(c_1 \oplus c_2)}       & (\inl{v})   & = & \inl{(\denot{c_1}~v)}               \\
        \denotbox{(c_1 \oplus c_2)}       & (\inr{v})   & = & \inr{(\denot{c_2}~v)}               \\
        \denotbox{(c_1 \otimes c_2)}      & (v_1 , v_2) & = & (\denot{c_1}~v_1 , \denot{c_2}~v_2)
      \end{tabular}}}
\end{center}

@@

\begin{theoremrep}\label{thm:semone}
  The semantics is sound in the following sense:
  \begin{itemize}
    \item For every level-1 combinator $c : A \isoone B$, we have a bijection $\denot{c}$ between $\denot{A}$ and $\denot{B}$.
    \item For every pair of combinators $c_1$ and $c_2$ of the same type $A \isoone B$, if there exists a level-2
          combinator $\alpha : c_1 \isotwo c_2$, then $\denot{c_1} = \denot{c_2}$, using extensional equivalence of
          functions.
  \end{itemize}
\end{theoremrep}
\begin{proof}
  For every primitive combinator $c$ listed on one side of~\cref{fig:pi-terms}, let $!c$ be the combinator listed on
  the other side. Thus $! \assoclp$ is $\assocrp$ and $! \swapp$ is $\swapp$ itself. Then we have that $\denot{c}$ and
  $\denot{!c}$ form an equivalence. For the level-2 combinator \Afun{idr◎l}, we check
  $\denot{\AgdaBound{c}~\AgdaOperator{\AgdaInductiveConstructor{◎}}~\AgdaInductiveConstructor{id⟷₁}}
    = \mathit{id} \circ \denot{c} = \denot{c}$. The other cases are only slightly more involved calculations.
\end{proof}
